Foster Children and Adoptees in New York City by Vera Sticker
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(tidyr)
library(dplyr)
library(reshape)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
fosters <- read.csv('fosNY_c_f.csv')
adoptions <- read.csv('adoNY_c_f.csv')
```

```{r echo=FALSE, Clean_the_Data}
#allCases <- merge(fosters, adoptions, by.x="RecNum", by.y="RecNum", all=T)
```

##The Data

This project analyzes two datasets, [Dataset 200](https://www.ndacan.acf.hhs.gov/datasets/dataset-details.cfm?ID=200) and [https://www.ndacan.acf.hhs.gov/datasets/dataset-details.cfm?ID=201](Dataset
201), provided by the Adoption and Foster Care Analysis and Reporting System 
(AFCARS), a federally-mandated data collection system which states use to report 
adoptions and foster care placements. These datasets were modified (reduced
variables and tidied) and uploaded by [Data4Democracy on Data.World](https://data.world/data4democracy/foster-care-nyc).

In this analysis, I want to explore this data to glean a better understanding 
of foster placements and adoptions in New York. Specifically, I am curious as to 
whether certain factors lead to more successful adoptions vs. foster care 
placements. 

To prepare the data for analysis, I rename several columns for intelligibility
and internal consistency and factorized several variables. This leads to the 
following information in the foster children dataset:

```{r echo=FALSE}
str(fosters)
```

"Fosters" contains 19 variables and 2377 observations.

The adoptions dataset:

```{r echo=FALSE}
str(adoptions)
```

"Adoptions" contains 13 variables and 1933 observations.

I will further explain the variable names as they are explored below. 

# Univariate Plots Section

I want to begin with the "fosters" data, since this is 
the child pool from which adoptions are typically chosen. I
want to know more about the characteristics of foster children as a whole.
The first relevant piece of information that comes to mind is age. 

```{r echo=FALSE, Univariate_Plots}

ggplot(aes(x=AgeAtEnd), data=fosters) +
  geom_histogram(binwidth=1)

#allCasesAges <- melt(subset(allCases, Adopted=="0")[, c("RecNum", "AgeAtEnd", "AgeAtAdopt")], id.vars='RecNum')
#ggplot(data=allCasesAges) +
#  geom_histogram(aes(x=value, (..count..)/sum(..count..), fill=variable), 
#                 alpha=0.3, binwidth=1, position="identity")
```
##Age




# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!