Foster Children and Adoptees in New York City by Vera Sticker
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(tidyr)
library(dplyr)
library(reshape)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
fosters <- read.csv('fosNY_c_f.csv')
adoptions <- read.csv('adoNY_c_f.csv')
```

```{r echo=FALSE, Clean_the_Data}
#allCases <- merge(fosters, adoptions, by.x="RecNum", by.y="RecNum", all=T)
```

##The Data

This project analyzes [Foster Care File 2015](https://www.ndacan.acf.hhs.gov/datasets/dataset-details.cfm?ID=200) from
the Adoption and Foster Care Analysis and Reporting System (AFCARS), a 
federally-mandated data collection system which states use to report 
adoptions and foster care placements. This dataset was tidied and uploaded by [Data4Democracy on Data.World](https://data.world/data4democracy/foster-care-nyc).

In this analysis, I want to explore this data to glean a better understanding 
of foster placements and adoptions in New York. Specifically, I am curious as to 
whether certain factors lead to more successful adoptions vs. foster care 
placements. 

To prepare the data for analysis, I rename several columns for intelligibility
and internal consistency and factorized several variables. This leads to the 
following information:

```{r echo=FALSE}
str(fosters)
```

"Fosters" contains 19 variables and 2377 observations.

```{r echo=FALSE}
summary(fosters)
```

"Adoptions" contains 13 variables and 1933 observations.

I will further explain the variable names as they are explored below. 

# Univariate Plots Section

I want to begin with the "fosters" data, since this is the child pool from 
which adoptions are typically chosen. I want to know more about the 
characteristics of foster children as a whole. The first relevant piece of 
information that comes to mind is age. 

```{r echo=FALSE, Univariate_Plots}

ggplot(aes(x=AgeAtEnd), data=fosters) +
  geom_histogram(binwidth=1)

#allCasesAges <- melt(subset(allCases, Adopted=="0")[, c("RecNum", "AgeAtEnd", "AgeAtAdopt")], id.vars='RecNum')
#ggplot(data=allCasesAges) +
#  geom_histogram(aes(x=value, (..count..)/sum(..count..), fill=variable), 
#                 alpha=0.3, binwidth=1, position="identity")
```
##Age

"AgeAtEnd" contains a foster child's age either at their time of adoption (if 
adopted) or at the end of the fiscal year (if not). This histogram shows all
foster children, both adopted and not. It seems to have a normal-ish 
distribution around age 14, but is slightly skewed towards the younger side. 

Next, I want to look at categorical variables such as sex, disability status,
and race of child.

```{r echo=FALSE}
ggplot(aes(x=Sex), data=fosters) +
  geom_bar()

```

```{r echo=FALSE}

ggplot(aes(x=ClinDis), data=fosters) +
  geom_bar()

```
```{r echo=FALSE}

ggplot(aes(x=ChildRace), data=fosters) +
  geom_bar()

```

```{r echo=FALSE}
ggplot(aes(x=Transracial), data=fosters) +
  geom_bar()
```

##Sex

Sex is a factor with two levels: 1 for male and 2 for female. There appear to be
slightly more males in the fosters data. 

##Disability

"ClinDis," standing for Clinical Disability, is a factor with two levels: 1 for
a child with a diagnosed disability, 2 for no diagnosis. An NA here means
disability status is unknown. Surprisingly, there are over twice as many foster
children with disabilities as without. 

##Race

Race is a factor with seven levels, each number corresponding to a 
race/ethnicity: 1 - White, 2 - Black, 3 - American Indian, 4 - Asian, 5 - 
Hawaiian/Pacific Islander, 6 - More than One Race, 7 - Hispanic, 99 - Unknown.
The majority of foster children are black, then white, then Hispanic.

##Transracial Placements
I also want to see if the child's race differs from the foster family's race.
"Transracial" is a factor with five levels, each number corresponding to a
status: 0 - not transracial, 1 - white parent/non-white child, 2 - non-white
parent/white child, 3 - other transracial placement, 9 - unknown, NA - missing.
Most known cases appear to be same-race placements. 

There is one more set of variables I am interested in: those that relate to a 
child's status in the foster system. 

```{r echo=FALSE}
ggplot(aes(x=TotalRem), data=fosters) +
  geom_histogram(binwidth=0.5)
```

```{r echo=FALSE}

ggplot(aes(x=PlacesNumber), data=fosters) +
  geom_histogram(binwidth=1)
```
```{r echo=FALSE}
ggplot(aes(x=MannerOfRem), data=fosters) +
  geom_bar()
```
```{r echo=FALSE}
ggplot(aes(x=Adopted), data=fosters) +
  geom_bar()
```

##Total Removals
"TotalRem" refers to the number of times a child has been removed from their
birth home, including the current time. The majority of children in the data 
appear to have been removed only once.

##Number of Placements
"PlacesNumber" refers to the number of places a child has lived, including their
current foster home, during the current removal episode (i.e. since leaving
their birth home the latest time). This is heavily negative-skewed, but has a 
long tail towards the right. Some children have been in 20 or more homes!

##Manner of Removal
"MannerOfRem" refers to how a child ended up in the foster system: a voluntary
surrender, a court order, or unknown."NA" refers to missing values in this case.
The vast majority of children were removed by court order.

##Adoptions
Finally, "Adopted" is a factor with two levels: 0 for not adopted, 1 for 
adopted. About 1/4 of the children in the dataset have been adopted. In the
bivariate analysis, we can explore this, and the adoptions dataset, to see what
factors may lead to this.

# Univariate Analysis

### What is the structure of your dataset?

My dataset is a dataframe containing 19 variables and 2377 observations. 12
variables are categorical (ordinal) and 6 are numeric (4 continuous, 2 discrete);
the remaining variable is RecNum which contains primary key information.

### What is/are the main feature(s) of interest in your dataset?

The feature of most interest is the "adopted" variable, which indicates whether
a child was adopted or not. About 1/4 of the children in the dataset were
adopted. I think that most of the bivariate and multivariate analyses I run will
contain this feature.

Based on the above plots, I would say that child characteristics such as age,
sex, race, and disability status will be of interest. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I am still interested in doing some analysis on number of removals and number
of places per removal as it relates to adoptions. I am wondering if children 
who go through a number of removals, or a number of placements, are more or less
likely to be adopted, possibly as a multivariate analysis with some of the child
characteristics.

### Did you create any new variables from existing variables in the dataset?

I did not create new variables for this project. I did rename most of the
existing variables for easier intelligibility and factorized the categorical 
variables. 

### Of the features you investigated, were there any unusual distributions?

I was surprised by the distribution of "ClinDis", which shows a nearly three-to-
one ratio of children with disabilities to children without. I was also surprised
by the long tail of "PlacesNumber" and, to a lesser extent, "TotalRem".


# Bivariate Plots Section


```{r echo=FALSE, Bivariate_Plots}
ggplot(aes(x=Adopted, y=AgeAtEnd), data=fosters) +
  geom_boxplot()
```
```{r echo=FALSE}
ggplot(aes(x=Adopted), 
       data=subset(fosters, !is.na(ClinDis))) +
       geom_bar(aes(fill=ClinDis), position='fill')
```
```{r echo=FALSE}
ggplot(aes(x=Adopted), 
       data=subset(fosters, !is.na(Sex))) +
       geom_bar(aes(fill=Sex), position='fill')
```
```{r echo=FALSE}
ggplot(aes(x=Adopted), 
       data=subset(fosters, !is.na(ChildRace))) +
       geom_bar(aes(fill=ChildRace), position='fill')

```
```{r}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!